Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-06-29 manuscript/manuscript.tex
Working to summarize paper to fewer pages
Small corrections to the R code (2 programs) to improve graphics and address color blind and B&W colors.  Manuscript review to reduce document length is about 1/4 complete

\usepackage[margin = 1in]{geometry}
\usepackage{amsmath}
\allowdisplaybreaks
\usepackage{authblk}
\usepackage{setspace}
\usepackage{natbib}
\usepackage{hyperref}\hypersetup{colorlinks=true, citecolor=blue}
% \usepackage{parskip}
\usepackage{textgreek}
\usepackage{graphicx}
\usepackage{booktabs} % For better looking tables
\usepackage{siunitx} % For alignment of numbers
\usepackage{placeins}
\usepackage{caption}  % For custom captions
\usepackage{float}   % Required for the H specifier
\usepackage{adjustbox}
\usepackage[]{lineno}
\linenumbers*[1]

@@

\jy{cut down to 6-8}
\noindent
\textit{Keywords:}
	311 Service Request Data;
	Data cleansing;
	Data Consistency;
	Data Curation;
	Data Democratization;
	Data Efficiency;
	Data science;
	Data Validity;
	Government Data;
	NYC Open Data;
	Open Data Movement;
	Open data;
	Public Engagement;
	Quality control;
	Research Data Management;
	Smart Cities;
	Transparency;
\end{abstract}

@@

\dt{
A note on naming conventions: dataset field names are named 
using ``snake\_case'' where each word in the variable name 
is written in lowercase, and words are separated by 
underscores, e.g. created\_date, complaint\_type, zip\_code, etc..   
}

\section{NYC 311 Service Request Data} 
\label{sec:data}

The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
record high of 348,463 monthly service requests in August 2020 due to
the COVID-19 pandemic, the 2021 expansion to include the MTA's city
subway system, and a record 3.23 million service requests in
2023. Today, the NYC 311 data system manages over 3 million service
requests per year, covering issues such as street noise, illegal
parking, heat and hot water, abandoned vehicles, and unsanitary
conditions. This data is publicly accessible through the NYC Open Data
Portal, which provides tools for querying, grouping, aggregating,
geo-mapping, visualizing, and exporting results.

@@

\jy{FixME: this paragraph's location will be revisited later.}
We extracted 311 SR data for the period of CY 2022--2023, a dataset of
6.4 million SR records. This is a large dataset which enables 
observation of some rarely seen events, data inaccuracies, 
and logical inconsistencies. It provides enough data to 
examine trends and data outlier analysis. We queried the data 
directly from the NYC Open Data Portal and exported the 
data for analysis with custom written R programs.

\jy{Every float (figure/table) needs to be referred to in the text.}

\begin{figure}[tbp]
	\centering
  	\includegraphics[width=0.8\textwidth]{10-year-trend-monthly.pdf}
  	\caption{10-year (2014-2023) Monthly SR Counts}
  	\label{fig:10-yr-monthly}
\end{figure}

We also looked at a 10-year period (2014-2023) to
observe system performance over a longer time-frame; see
Figure~\ref{fig:10-yr-monthly}. During that time-frame,
the number of 311 SRs grew by 50\%. The summer and early fall months 
make up all of the Top 5 SR counts by day.

\begin{table}[tbp]
    \centering
    \caption{Top 5 and Bottom 5 Days for SRs}
    \label{tab:top-10-counts}
    \begin{tabular}{@{}lll lll@{}}
      \toprule
      \multicolumn{2}{c}{Top 5 Days} & \multicolumn{2}{c}{Bottom 5 Days} \\
      \cmidrule(r){1-2} \cmidrule(l){3-4}
      Date & Count & Date & Count\\
      \midrule
      2020-08-04 & 24,415 & 2014-12-25 & 2965\\
      2020-08-05 & 19,560 & 2015-12-25 & 3247\\
      2023-09-29 & 17,962 & 2014-04-20 & 3287\\
      2020-07-05 & 16,916 & 2015-12-26 & 3405\\
      2020-06-21 & 15,883 & 2016-11-06 & 3419\\
      \bottomrule
    \end{tabular}
\end{table}

Table~\ref{tab:top-10-counts} summarizes the top 5 and bottom 5 days
in counts of SRs. The highest daily SR count, 24,415, occurred on
August 4, 2020, coinciding with the aftermath of Tropical Storm
Isaias, highlighting the system's critical role during crises. The
second-highest count, 19,560, was the following day, reflecting the
storm's continued impact. 
Conversely, the lowest daily SR count, 2,965, was on Christmas Day,
2014, with similar lows on other Christmas Days, indicating reduced
activity during major holidays. The data also shows fewer requests on
weekends, such as April 20, 2014 (Easter Sunday), and November 6, 2016
(a non-holiday Sunday).

\begin{figure}[tbp]
	\centering
	\includegraphics[width=0.8\textwidth]{SRs_by_Agency.pdf}
  	\caption{SR counts by Agency with Cumulative Percentage}
	\label{fig:SRcountbyAgency}
\end{figure}

Figure~\ref{fig:SRcountbyAgency} display the SR counts by agency with
cumulative percentage. Two important fields in this analysis are the
responsible NYC Agency and the type of complaint. Given the size of
the City of New York government, identifying the responsible agency
and the type of complaint is essential for pinpointing the accountable
party and troubleshooting discrepancies. This identification helps
determine whether an issue is specific to an agency or indicative of a
systemic problem. Although 311 SRs involve 16 different agencies, six
agencies account for 90\% of the complaints. These are the New York
Police Department (NYPD), Housing Preservation and Development (HPD),
New York City Department of Sanitation (DSNY), Department of
Transportation (DOT), Department of Environmental Protection (DEP),
and Department of Parks and Recreation (DPR).

\begin{figure}[tbp]
  \centering
  \includegraphics[width=0.8\textwidth]{SR_by_Complaint_Type.pdf} 
  \caption{Top 20 complaint\_type(s) and Cumulative Percentage} 
  \label{fig:SR_complaints}
\end{figure}

\begin{table}[tbp]
  \centering
  \caption{Noise-related complaints\_type(s) by count with Agency}
  \label{tab:noisecomplaints}
  \begin{tabular}{@{}lS[table-format=7.0,round-mode=places,
    round-precision=0]S[table-format=2.2,round-mode=places,
    round-precision=2]l@{}} % 'l' for left-aligned, 'S' for siunitx number alignment
    \toprule
    \textbf{complaint\_type} & \textbf{Count} & \textbf{Percentage} & \textbf{Agency} \\ 
    \midrule
    Noise - Residential        & 675502 & 10.56 & NYPD  \\ 
    Noise - Street/Sidewalk    & 300507 &  4.70 & NYPD  \\ 
    Noise - Commercial         & 128892 &  2.02 & NYPD  \\ 
    Noise - Vehicle            & 115956 &  1.81 & NYPD  \\ 
    Noise                      & 100413 &  1.57 & DEP   \\ 
    Noise - Helicopter         &  85345 &  1.33 & EDC   \\ 
    Noise - Park               &  16620 &  0.26 & NYPD  \\ 
    Noise - House of Worship &   2757 &  0.04 & NYPD  \\ 
    \bottomrule
  \end{tabular}
\end{table}

The distribution of complaint types skews toward a few select
categories, accompanied by a long tail. Figure~\ref{fig:SR_complaints}
shows that the top 20 complaint types account for 70\% of the total
220 different types. Notably, these top complaints include several
variations of noise-related issues, which collectively represent 22\%
of all complaints. When consolidated, noise complaints emerge as the
most frequently occurring type. Table~\ref{tab:noisecomplaints}
provides a detailed breakdown of noise-related complaint types by
count and responsible agency. The most common noise complaint is
``Noise - Residential,'' with 675,502 complaints, accounting for 10.56\%
of all service requests, and primarily handled by the NYPD. Other
significant noise complaints include ``Noise - Street/Sidewalk"
(4.70\%), ``Noise - Commercial'' (2.02\%), and ``Noise - Vehicle"
(1.81\%), all managed by the NYPD. Additional categories such as
"Noise - Helicopter'' (1.33\%) fall under the jurisdiction of the EDC,
and ``Noise'' (1.57\%) is managed by the DEP. The prevalence of these
complaints underscores the critical role of the NYPD and other
agencies in addressing noise-related issues, reflecting the
significant impact of urban noise on residents' quality of life. This
data highlights the importance of efficient response and management
strategies to mitigate noise disturbances in New York City.

@@

\section{Structural Issues}
\label{sec:structural}
Structural issues in the context of data cleansing refers to issues 
related to the how data is organized, formatted, or structured within a 
dataset. Structural issues can make it difficult to analyze the 
data effectively. Here are some characteristics of the 311 SR data set:

@@

\textbf{Issue: Fields not in the Data Dictionary} The 311 SR 
\href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}
 identifies 41 data columns (fields) along with related information 
 for each column. However, when you download the data or 
 explore it via the online portal, you immediately note that there 
 are 47 columns (fields). The additional six fields are not addressed in the 
 Data Dictionary, but do and show up on the Column Manager widget
 as ``@computed\_region\_xxxx\_xxxx''. To some extent one can infer 
 what the these @computed\_fields are and how they are determined, but 
 it is not possible to know for certain. Most importantly, can these fields 
 be reliably used for analytical purposes?  These six fields are:

@@

\section{Validating data types}
\label{sec:datatypes}
Fortunately, both the Data Dictionary and the portal column manager 
do a good job of indicating the data type. The following fields were 
checked for their specified data types:

\begin{itemize}
	\item created\_date, closed\_date, due\_date, and resolution\_action\_updated\_date: 
	All are all in proper date format.
	\item zip\_codes and incident\_zip: All are 5 numeric digits except 
	for two non-numeric entries for incident\_zip (``na'' and ``N/A'').
	\item x\_coordinate\_state\_plane \&  y\_coordinate\_state\_plane 
	(a NY State geo-location system): All are all numeric.
	\item latitude \& longitude: Both are numeric.
	\item community\_districts, borough\_boundaries, 
	city\_council\_district, police\_precinct, and police\_precincts: All are numeric.
	\item Free-form text fields that contain a ``,'' (e.g. incident\_address 
	\& resolution\_description) are properly enclosed in quotation marks.
\end{itemize}

\section{Blank \& N/A data}
\label{sec:blanks}
Understanding the absence of data by field is an important factor 
when undertaking analysis. For example, if you wanted to see if the SRs were
closed before or after their due\_date, you would be challenged 
as 99.57\% of the due\_date field is blank. Which calls into 
question...are these missing values, or is the concept of a 
due\_date simply not applicable for the vast majority of SRs?

When counting the various fields for blank or N/A values, the fields appear 
to divide into three groups: \textbf{Mostly Empty}, \textbf{Partially Empty}, 
and \textbf{Few/None Empty}. he Mostly Empty category ranges from 93-99.9\% blank. 
It includes such fields as taxi\_company due\_date, pickup\_location, and 
landmark. The Partially Empty includes such fields as location\_type, borough, 
and cross\_street. And the Few/None Empty includes created\_date, 
complaint\_type, agency, and status. In some cases it may make 
sense to inquiry as to why some fields are frequently blank. Here is a graphic 
depiction of total empty (blank \& N/As) for each fields. You can see 
the natural grouping into the Most, Some, and Few categories 
in the stair-type visualization.

\begin{figure}[tbp]
	\centering
  	\includegraphics[width=\textwidth]{BlankFields.pdf}
	\caption{Number and Percentage of Empty/Blank Entries}
	\label{fig:blank_fields}
\end{figure}

 \section{Validating Data for Acceptable Values}
 \label{sec:domain}
 In this study, testing for invalid values revealed several issues 
 associated with select fields. Accordingly, an analyst needs to be 
 extra diligent if using these fields for analysis, and in all likelihood, 
 rows with invalid field values would  need to be removed from the 
 analysis. Here are some results.

\begin{itemize}
	\item Latitude and Longitude fields fell within the geographic 
	boundaries of New York City. 
	\item The unique\_key field was in fact unique.
	\item Many fields have a domain of acceptable values, often 
	determined from common usage or by examining historical 
	datasets. Unfortunately, these domains of acceptable 
	values are not specified in the Data Dictionary. These fields all tested as 
	compliant within their domain:
	\begin{itemize}
		\item address\_type
		\item status
		\item borough, borough\_boundaries, \& park\_borough 
		\item data\_channel
		\item vehicle\_type
		\item city\_council\_district
	\end{itemize}	
\end{itemize}

\subsection{Issues with Zip Codes}
\label{sec:zipcodesissues}
 Unfortunately some fields proved to be  problematic when comparing 
 the values to a domain of legal values. For example, all zip codes 
 (two fields: zip\_codes and incident\_zip) should be valid as defined 
 by the USPS database which contains 37,946 valid zip codes. However,
 the computed field zip\_codes proved problematic with 
58\% (3.6 million) of the field entries being invalid. Even the 
field incident\_zip while having only .07\% invalid entries, exhibited
4163 such errors. The breakdown of the invalid entries in the zip\_code, sorted by 
Agency shows that the distribution by percentage almost precisely 
mirrors the overall breakdown of SRs by Agency, potentially indicating
a systemic problem.

\subsubsection{Case Study: Noise Complaints by Zip Code}
\label{sec:case-study-zip-codes}
\textbf{Scenario:} The NYC Office of Nightlife (ONL) wants to 
know ``What are the top 10 zip codes for Noise Complaints (all 8 types) 
over the last two years?'' The intent is to assess the impact of the 
recent NYC effort promoting a vibrant nightlife 
in NYC while seeking to ease strained relations 
between bar and club owners. Let's assume that this analysis 
uses the zip\_codes field, one of the (six) 
computed fields that has shown to have validity problems. 
We then repeat the analysis with the incident\_zip field instead. Let's 
subject these two data sets to validation against the USPS zip code database.

\begin{table}[tbp]
    \centering
    \caption{Comparison of Top Ten Zip Codes Lists}
	    \begin{tabular}{@{}lS[table-format=6.0,round-mode=places,
	    round-precision=0]c@{\hskip 0.5cm}@{}lS[table-format=6.0,
	    round-mode=places,round-precision=0]c@{}}
		\toprule
	 	\multicolumn{3}{c}{\textbf{zip\_codes}} & \multicolumn{3}{c}{\textbf{incident\_zip}} \\
	      \cmidrule(r){1-3} \cmidrule(l){4-6}
	      \textbf{Zip Code} & \textbf{Count} & \textbf{Valid?} 
	      & \textbf{Zip Code} & \textbf{Count} & \textbf{Valid?} \\
	      \midrule
	        11275 & 104556 & FALSE & 10466 & 104562 & TRUE \\
	        12420 & 27503 & TRUE & 10023 & 27972 & TRUE \\
	        12428 & 26564 & TRUE & 10031 & 25548 & TRUE \\
	        10935 & 25508 & FALSE & 10457 & 25066 & TRUE \\
	        10934 & 23448 & FALSE & 10453 & 24752 & TRUE \\
	        10931 & 22381 & TRUE & 10456 & 24751 & TRUE \\
	        10930 & 22121 & TRUE & 10452 & 22527 & TRUE \\
	        17613 & 21963 & FALSE & 10025 & 21705 & TRUE \\
	        10936 & 21707 & FALSE & 10458 & 21689 & TRUE \\
	        11606 & 21435 & FALSE & 10032 & 20622 & TRUE \\
	      \bottomrule
	    	\end{tabular}
 	\label{tab:zipcodes}
\end{table}

As indicated, six out of ten zip\_codes are invalid, which corresponds closely 
with what is observed in the overall dataset (58\%). Whereas the incident\_zip 
field is completely valid, again in-line with the overall incident\_zip 
validation percentage (99.04\%). If the analysis performed using the 
zip\_codes field were to be 	presented to the Director, ONL, the conclusion 
would be in error.

\subsection{Issues with Police Precincts} 	
\label{sec:police-precincts}
A curious case also exists when examining the two nearly identical 
fields - police\_precincts and police\_precinct. Both of those fields 
are among the ``computed'' fields in the dataset. Using 
\href{https://www.nyc.gov/site/nypd/bureaus/patrol/precincts-landing.page}
{NYPD Precinct listings} it's possible to determine the valid 
police precincts; they are generally numeric or have numeric 
representations in the dataset. However, what we find is that 
both fields police\_precinct and police\_precincts  have  35\% 
invalid entries. Unfortunately, it's not the same invalid entries. For 
the police\_precincts field, there are 2,171,864 invalid entries 
(35\%)., a consequential number of errors. Similarly for the 
police\_precinct field, there are 2,171,778 invalid entries 
(also 35\%).

\subsection{Issues with Community Boards}
\label{sec:communityboards}
Community Boards are an important aspect of NYC government. They are the 
most local, grassroots form of City government, and a vital connection 
between communities and elected officials and City agencies. Community 
Boards are used as a measure of City services throughout the five 
boroughs. In the 2022-2023 dataset there are 27,276 invalid community\_board 
entries which represents 0.43\% of non-blank data. The distribution 
of invalid Community Boards by Agency is not consistent with 
the overall SR Agency distribution. This indicates that there are likely 
specific issues at some key Agencies, in this case Taxi \& Limo 
Commission (TLC), Parks \& Recreation, etc.

\subsection{Issues with Community Districts}
\label{sec:communitydistrict}
Another of the ``computed'' fields is community\_districts. Community Districts 
are the boundaries for the Community Boards, but unlike how 
Community Boards are an arm of the government of New York City, 
the Community District is used by the Department of City Planning 
(DCP) for purposes of environmental, socio-economic, and demographic 
purposes. It is a geographical division rather than a local 
government division. Due to how the community\_district data 
is formatted, it is not possible to establish validity. However, 
it is possible to determine that the dataset contains 72 unique 
entries, while there are only 59 valid Community Districts.

\subsection{created\_date and closed\_date(s) -- Negative Duration}
\label{sec:negativeduration}
For the created\_date and closed\_date fields, you might 
expect that these two fields are automatically populated by the 
SR application software. Thus the created\_date is associated with setting an SR status 
to ``new'' or ``open''. And similarly when an SR's 	status is 
changed to ``closed''  the closed\_date would be 
populated. Unfortunately, that does not appear to be the case as
there are several anomalies with the various date fields.

>>>>

\usepackage[margin = 1in]{geometry}
\usepackage{amsmath}
\allowdisplaybreaks
\usepackage{authblk}
\usepackage{setspace}
\usepackage{natbib}
\usepackage{hyperref}\hypersetup{colorlinks=true, citecolor=blue}
% \usepackage{parskip}
\usepackage{textgreek}
\usepackage{graphicx}
\usepackage{booktabs} % For better looking tables
\usepackage{siunitx} % For alignment of numbers
\usepackage{placeins}
\usepackage{caption}  % For custom captions
\usepackage{float}   % Required for the H specifier
\usepackage{adjustbox}
\usepackage[]{lineno}
\usepackage{comment}
\linenumbers*[1]

@@

\noindent
\textit{Keywords:}
	Data cleansing;
	Data Curation;
	Data science;
	NYC Open Data;
	Open data;
	Smart Cities;
	Transparency;
\end{abstract}

@@

\section{NYC 311 Service Request Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
record high of 348,463 monthly service requests in August 2020 due to
the COVID-19 pandemic, the 2021 expansion to include the MTA's city
subway system, and a record 3.23 million service requests in
2023. Today, the NYC 311 data system manages over 3 million service
requests per year, covering issues such as street noise, illegal
parking, heat and hot water, abandoned vehicles, and unsanitary
conditions. This data is publicly accessible through the NYC Open Data
Portal, which provides tools for querying, grouping, aggregating,
geo-mapping, visualizing, and exporting results.

@@

\begin{comment}
\jy{FixME: this paragraph's location will be revisited later.}
We extracted 311 SR data for the period of CY 2022--2023, a dataset of
6.4 million SR records. This is a large dataset which enables 
observation of some rarely seen events, data inaccuracies, 
and logical inconsistencies. It provides enough data to 
examine trends and data outlier analysis. We queried the data 
directly from the NYC Open Data Portal and exported the 
data for analysis with custom written R programs.
\end{comment}

%\begin{figure}[tbp]
%	\centering
 % 	\includegraphics[width=0.8\textwidth]{10-year-trend-monthly.pdf}
 %	\caption{10-year (2014-2023) Monthly SR Counts}
 % 	\label{fig:10-yr-monthly}
%\end{figure}

To begin, we looked at a 10-year period (2014-2023) to
observe the data over a longer time-frame; see
Figure~\ref{fig:10-yr-monthly}. During that time-frame,
the number of 311 SRs grew by nearly 50\%. The summer months make
up the highest days, and the winter months show the lowest activity.

\begin{comment}
\begin{table}[tbp]
    \centering
    \caption{Top 5 and Bottom 5 Days for SRs}
    \label{tab:top-10-counts}
    \begin{tabular}{@{}lll lll@{}}
      \toprule
      \multicolumn{2}{c}{Top 5 Days} & \multicolumn{2}{c}{Bottom 5 Days} \\
      \cmidrule(r){1-2} \cmidrule(l){3-4}
      Date & Count & Date & Count\\
      \midrule
      2020-08-04 & 24,415 & 2014-12-25 & 2965\\
      2020-08-05 & 19,560 & 2015-12-25 & 3247\\
      2023-09-29 & 17,962 & 2014-04-20 & 3287\\
      2020-07-05 & 16,916 & 2015-12-26 & 3405\\
      2020-06-21 & 15,883 & 2016-11-06 & 3419\\
      \bottomrule
    \end{tabular}
\end{table}
\end{comment}

%Table~\ref{tab:top-10-counts} summarizes the top 5 and bottom 5 days
i%n counts of SRs.

The highest daily SR count, 24,415, occurred on
August 4, 2020, coinciding with the aftermath of Tropical Storm
Isaias, highlighting the system's critical role during crises.

% The second-highest count of 19,560 was the following day, 
%reflecting the storm's continued impact.

Conversely, the lowest daily SR count, 2,965, was on Christmas Day,
2014, with similar lows on other Christmas Days, indicating reduced
activity during major holidays.

% The data also shows fewer requests on weekends, such as 
%April 20, 2014 (Easter Sunday), and November 6, 2016 (a non-holiday Sunday).

%\begin{figure}[tbp]
%	\centering
%	\includegraphics[width=0.8\textwidth]{SRs_by_Agency.pdf}
 % 	\caption{SR counts by Agency with Cumulative Percentage}
%	\label{fig:SRcountbyAgency}
%\end{figure}

Figure~\ref{fig:SRcountbyAgency} display the SR counts by agency with
cumulative percentage. Two important fields in this analysis are the
responsible NYC Agency and the type of complaint. Identifying the 
responsible agency and type of complaint is essential 
for identifying the accountable party and troubleshooting discrepancies. 
This identification also assists in determining whether an issue is 
specific to an agency or indicative of a broader, systemic problem. 
Six agencies account for 90\% of the complaints. These are the New York
Police Department (NYPD), Housing Preservation and Development (HPD),
New York City Department of Sanitation (DSNY), Department of
Transportation (DOT), Department of Environmental Protection (DEP),
and Department of Parks and Recreation (DPR).

\begin{comment}
\begin{figure}[tbp]
  \centering
  \includegraphics[width=0.8\textwidth]{SR_by_Complaint_Type.pdf} 
  \caption{Top 20 complaint\_type(s) and Cumulative Percentage} 
  \label{fig:SR_complaints}
\end{figure}

\begin{table}[tbp]
  \centering
  \caption{Noise-related complaints\_type(s) by count with Agency}
  \label{tab:noisecomplaints}
  \begin{tabular}{@{}lS[table-format=7.0,round-mode=places,
    round-precision=0]S[table-format=2.2,round-mode=places,
    round-precision=2]l@{}} % 'l' for left-aligned, 'S' for siunitx number alignment
    \toprule
    \textbf{complaint\_type} & \textbf{Count} & \textbf{Percentage} & \textbf{Agency} \\ 
    \midrule
    Noise - Residential        & 675502 & 10.56 & NYPD  \\ 
    Noise - Street/Sidewalk    & 300507 &  4.70 & NYPD  \\ 
    Noise - Commercial         & 128892 &  2.02 & NYPD  \\ 
    Noise - Vehicle            & 115956 &  1.81 & NYPD  \\ 
    Noise                      & 100413 &  1.57 & DEP   \\ 
    Noise - Helicopter         &  85345 &  1.33 & EDC   \\ 
    Noise - Park               &  16620 &  0.26 & NYPD  \\ 
    Noise - House of Worship &   2757 &  0.04 & NYPD  \\ 
    \bottomrule
  \end{tabular}
\end{table}
\end{comment}

The distribution of complaint types skews toward a few select
categories, accompanied by a long tail. Figure~\ref{fig:SR_complaints}
shows that the top 20 complaint types account for 70\% of the total
220 different types. Notably, these top complaints include several
variations of noise-related issues, which collectively represent 22\%
of all complaints. When consolidated, noise complaints emerge as the
most frequently occurring complaint type.

% Table~\ref{tab:noisecomplaints}
%provides a detailed breakdown of noise-related complaint types by
%count and responsible agency. The most common noise complaint is
%``Noise - Residential,'' with 675,502 complaints, accounting for 10.56\%
%of all service requests, and primarily handled by the NYPD. Other
%significant noise complaints include ``Noise - Street/Sidewalk"
(%4.70\%), ``Noise - Commercial'' (2.02\%), and ``Noise - Vehicle"
(%1.81\%), all managed by the NYPD. Additional categories such as
%"Noise - Helicopter'' (1.33\%) fall under the jurisdiction of the EDC,
%and ``Noise'' (1.57\%) is managed by the DEP. The prevalence of these
%complaints underscores the critical role of the NYPD and other
%agencies in addressing noise-related issues, reflecting the
%significant impact of urban noise on residents' quality of life. This
%data highlights the importance of efficient response and management
%strategies to mitigate noise disturbances in New York City.

@@

\section{Structural Issues}
\label{sec:structural}
Structural issues refers to how data is organized, formatted, 
or structured within a dataset. Structural issues can make 
it difficult to analyze the data effectively. Here are some 
characteristics of the 311 SR data set:

@@

\textbf{Issue: Fields not in the Data Dictionary} The 311 SR 
\href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}
 identifies 41 data columns (fields) along with related information 
 for each column. However, when downloaded the data contains 
 six fields are not included in the Data Dictionary. However, these
 fields do show up on the portal Column Manager widget
 as ``@computed\_region\_xxxx\_xxxx''. To some extent one can infer 
 what the these @computed\_fields are and how they are determined but
 not with certainty. Most importantly, can these fields be reliably 
 used for analytical purposes?  These six fields are:

@@

\section{Validating data types}
\label{sec:datatypes}
Fortunately, both the Data Dictionary and the portal column manager 
clearly indicate the epected data type. The following fields were 
checked for their specified data types:

\begin{itemize}
	\item created\_date, closed\_date, due\_date, and resolution\_action\_updated\_date: 
	All are all in proper date format.

	\item zip\_codes and incident\_zip: All are 5 numeric digits except 
	for two non-numeric entries for incident\_zip (``na'' and ``N/A'').

	\item x\_coordinate\_state\_plane \&  y\_coordinate\_state\_plane 
	(a NY State geo-location system): All are all numeric.

	\item latitude \& longitude: Both are numeric.

	\item community\_districts, borough\_boundaries, 
	city\_council\_district, police\_precinct, and police\_precincts: All are numeric.

	\item Free-form text fields that contain a ``,'' (e.g. incident\_address 
	\& resolution\_description) are properly enclosed in quotation marks.
\end{itemize}

\section{Blank \& N/A data}
\label{sec:blanks}
Understanding the absence of data by field is important 
when undertaking analysis. For example, if you wanted to see if the SRs were
closed before or after their due\_date, you would be challenged 
as 99.57\% of the due\_date field is blank. When counting 
the various fields for blank or N/A values, the fields appear 
to divide into three groups: \textbf{Mostly Empty}, \textbf{Partially Empty}, 
and \textbf{Few/None Empty}. The Mostly Empty category 
ranges from 93-99.9\% blank and includes such fields as 
taxi\_company due\_date, pickup\_location, and 
landmark. The Partially Empty includes such fields as location\_type, borough, 
and cross\_street. And the Few/None Empty includes created\_date, 
complaint\_type, agency, and status. In some cases it may make 
sense to inquiry as to why some fields are frequently blank. Here is a graphic 
depiction of total empty (blank \& N/As) for each fields illustrating
the grouping into the Most, Some, and Few categories.

%\begin{figure}[tbp]
%	\centering
 % 	\includegraphics[width=\textwidth]{BlankFields.pdf}
%	\caption{Number and Percentage of Empty/Blank Entries}
%	\label{fig:blank_fields}
%\end{figure}

 \section{Validating Data for Acceptable Values}
 \label{sec:domain}
Any analytic effort needs to be aware of fields which 
contain invalid values. Typically such fields would need to be
identified and removed from the analysis. Here are some results
of field validation.

\begin{itemize}
	\item Latitude and Longitude fields fell within the geographic 
	boundaries of New York City.

	\item The unique\_key field was in fact unique.

	\item Unfortunately, very few domains of acceptable 
	values are specified in the Data Dictionary. The below 
	fields were tested as complying with their domain as 
	determined by public usage and historical datasets:
	\begin{itemize}
		\item address\_type
		\item status
		\item borough, borough\_boundaries, \& park\_borough 
		\item data\_channel
		\item vehicle\_type
		\item city\_council\_district
	\end{itemize}	
\end{itemize}

%\subsection{Issues with Zip Codes}
\label{sec:zipcodesissues}
 \textbf{Issues with Zip Codes} Some fields proved to be problematic 
 as to complying with a domain of legal values. For example, all zip codes 
 (two fields: zip\_codes and incident\_zip) should be valid as defined 
 by the USPS database which contains 37,946 valid zip codes. However,
 the computed field zip\_codes proved problematic with 
58\% (3.6 million) of the field entries being invalid. And the 
incident\_zip field, while having only .07\% invalid entries, still contains
4163 such errors. The breakdown of the invalid entries in the zip\_code, sorted by 
Agency shows that the distribution by percentage mirrors the overall 
breakdown of SRs by Agency, potentially indicating a systemic problem.

%\subsubsection{Case Study: Noise Complaints by Zip Code}
\label{sec:case-study-zip-codes}
\textbf{Case Study: Noise Complaints by Zip Code} Scenario: The NYC 
Office of Nightlife (ONL) wants to know ``What are the top 10 
zip codes for Noise Complaints over the last two years?'' ONL wants
to assess the impact of the recent NYC effort promoting a vibrant nightlife 
in NYC while seeking to ease strained relations between bar 
and club owners. Let's assume that this analysis 
uses the zip\_codes field, one of the (six) computed fields known to 
have validity problems. We then repeat the analysis with the more 
accurate incident\_zip field. We subject these two analysis effort 
to validation against the USPS zip code database.

\begin{comment}	 
\begin{table}[tbp]
    \centering
    \caption{Comparison of Top Ten Zip Codes Lists}
	    \begin{tabular}{@{}lS[table-format=6.0,round-mode=places,
	    round-precision=0]c@{\hskip 0.5cm}@{}lS[table-format=6.0,
	    round-mode=places,round-precision=0]c@{}}
		\toprule
	 	\multicolumn{3}{c}{\textbf{zip\_codes}} & \multicolumn{3}{c}{\textbf{incident\_zip}} \\
	      \cmidrule(r){1-3} \cmidrule(l){4-6}
	      \textbf{Zip Code} & \textbf{Count} & \textbf{Valid?} 
	      & \textbf{Zip Code} & \textbf{Count} & \textbf{Valid?} \\
	      \midrule
	        11275 & 104556 & FALSE & 10466 & 104562 & TRUE \\
	        12420 & 27503 & TRUE & 10023 & 27972 & TRUE \\
	        12428 & 26564 & TRUE & 10031 & 25548 & TRUE \\
	        10935 & 25508 & FALSE & 10457 & 25066 & TRUE \\
	        10934 & 23448 & FALSE & 10453 & 24752 & TRUE \\
	        10931 & 22381 & TRUE & 10456 & 24751 & TRUE \\
	        10930 & 22121 & TRUE & 10452 & 22527 & TRUE \\
	        17613 & 21963 & FALSE & 10025 & 21705 & TRUE \\
	        10936 & 21707 & FALSE & 10458 & 21689 & TRUE \\
	        11606 & 21435 & FALSE & 10032 & 20622 & TRUE \\
	      \bottomrule
	    	\end{tabular}
 	\label{tab:zipcodes}
\end{table}
\end{comment}

As indicated, six out of ten zip\_codes are invalid, which corresponds closely 
with what is observed in the overall dataset (58\% invalid). Whereas 
the incident\_zip field is completely valid, in-line with the 
overall incident\_zip validation percentage of 99.04\%.

%\subsection{Issues with Police Precincts} 	
\label{sec:police-precincts}
A curious case exists when examining two (nearly) identical 
fields: police\_precincts and police\_precinct. Both of those fields 
are among the ``computed'' fields previously mentioned. Using 
\href{https://www.nyc.gov/site/nypd/bureaus/patrol/precincts-landing.page}
{NYPD Precinct listings} it is possible to determine what are valid 
police precincts. What we find is that both police\_precinct 
and police\_precincts  have  35\% invalid entries. Unfortunately, 
it's not the same invalid entries. For the police\_precincts field, 
there are 2,171,864 invalid entries (35\%). Similarly for the 
police\_precinct field, there are 2,171,778 invalid entries 
(also 35\%), but not the same invalid entries.

==========REVIEW BELOW HERE===========

%\subsection{Issues with Community Boards}
\label{sec:communityboards}
Community Boards are an important aspect of NYC government. They are the 
most local, grassroots form of City government, and a vital connection 
between communities and elected officials and City agencies. Community 
Boards are used as a measure of City services throughout the five 
boroughs. In the 2022-2023 dataset there are 27,276 invalid community\_board 
entries which represents 0.43\% of non-blank data. The distribution 
of invalid Community Boards by Agency is not consistent with 
the overall SR Agency distribution. This indicates that there are likely 
specific issues at some key Agencies, in this case Taxi \& Limo 
Commission (TLC), Parks \& Recreation, etc.

%\subsection{Issues with Community Districts}
\label{sec:communitydistrict}
Another of the ``computed'' fields is community\_districts. Community Districts 
are the boundaries for the Community Boards, but unlike how 
Community Boards are an arm of the government of New York City, 
the Community District is used by the Department of City Planning 
(DCP) for purposes of environmental, socio-economic, and demographic 
purposes. It is a geographical division rather than a local 
government division. Due to how the community\_district data 
is formatted, it is not possible to establish validity. However, 
it is possible to determine that the dataset contains 72 unique 
entries, while there are only 59 valid Community Districts.

%\subsection{created\_date and closed\_date(s) -- Negative Duration}
\label{sec:negativeduration}
For the created\_date and closed\_date fields, you might 
expect that these two fields are automatically populated by the 
SR application software. Thus the created\_date is associated with setting an SR status 
to ``new'' or ``open''. And similarly when an SR's 	status is 
changed to ``closed''  the closed\_date would be 
populated. Unfortunately, that does not appear to be the case as
there are several anomalies with the various date fields.
